# Generalized linear models

```{r}
library(ggplot2)
```


## 

There are three components to a GLM:

Random component – refers to the probability distribution of the response variable (Y); e.g. binomial distribution for Y in the binary logistic regression.
Systematic component - refers to the explanatory variables (X1, X2, ... Xk) as a combination of linear predictors; e.g. β0 + β1x1 + β2x2 as we have seen in logistic regression.
link Function, η or g(μ) - specifies the link between random and systematic components. It says how the expected value of the response relates to the linear predictor of explanatory variables; e.g. η = logit(π) for logistic regression.

## Offset

rate = count / unit time

offset converts a count to rate

## Poisson regression

Poisson regression models are generalized linear models with the logarithm as the link function.

```{r}
week <- 1:7
amount_time <- c(0.85, 1.1, 0.85, 1.2, 0.95, 1.1, 0.9)
count <- c(90, 76, 37, 27, 19, 13, 9)
rate <- count / amount_time
df <- data.frame(week, amount_time, count, rate)
```


$$\log(\text{count}) = \beta_0 + \beta \cdot \text{week}$$

If we are fitting $\text{rate}$ instead of $\text{count}$:

$$\begin{align}
\log(\text{rate}) = \log\left(\frac{\text{count}}{\text{amount time}}\right) &= \beta_0 + \beta \cdot \text{week} \\
\Leftrightarrow \log(\text{count}) - \log(\text{amount time}) &= \beta_0 + \beta \cdot \text{week} \\
\Leftrightarrow \log(\text{count}) &= \beta_0 + \beta \cdot \text{week} + \log(\text{amount time})
\end{align}$$

The $\log(\text{amount time})$ is an offset.

```{r}
mod <- glm(count ~ week + offset(log(amount_time)), family = poisson)
mod
```

```{r}
pred <- predict(mod, type = "response")
ggplot(df, aes(x = week, y = count)) +
  geom_point() +
  geom_line(aes(y = pred)) +
  theme_classic()
```

